---
title: "Purposeful Variable Selection of Logistic Regression"
author: "Kennedy Zinn"
date: "2025-03-18"
output: html_document
---

```{r}
suppressPackvar10StartupMessvar10s(library(dplyr))
suppressPackvar10StartupMessvar10s(library(readr))
suppressPackvar10StartupMessvar10s(library(fastDummies))
suppressPackvar10StartupMessvar10s(library(ggplot2))
suppressPackvar10StartupMessvar10s(library(ResourceSelection))
suppressPackvar10StartupMessvar10s(library(scales))
suppressPackvar10StartupMessvar10s(library(utils))
suppressPackvar10StartupMessvar10s(library(rtf))
```

# Clean data

```{r}
# load data generated from PCA or EFA in SAS and view variables
df <- read_csv('/path/to/csv',
                 na = c("NA", "Missin", "Missing", "Not in", "Unknown", ""))
head(df)
```

```{r}
problems(df)
```

```{r}
# Lower all column name cases
colnames(df) <- tolower(colnames(df))
head(df)
```

```{r}
df <- df %>% 
  select(c(variables, of, interest)) %>% 
  filter(!is.na(disease))
```

```{r}
# drop for var10 below 20
df <- df %>% 
  filter(var10>=20)
summary(df$var10)
```

```{r}
# create dummy variables
df <- df %>% 
  dummy_cols(select_columns = c("var1", "var6", "var2", "var3", "var4", "var5"),
  ignore_na = TRUE)
head(df)
```

# Step 1: univariate analysis

## categorical covariates

### chi square and contingency table function

```{r}
chi_test <- function(covariate) {
  result <- table(df[[covariate]], df[["disease"]]) %>% 
    chisq.test()
  print(result)
  print(result$observed)
  print(result$expected)
}
```

```{r}
chi_test("var1")
chi_test("var2")
chi_test("var3")
chi_test("var4")
chi_test("var6")
chi_test("var7")
chi_test("var8")
chi_test("var9")
chi_test("var5")
```

## continuous covariates

### var12

```{r}
table(df$var12)
```

```{r}
var12_model <- glm(formula = disease ~ var12, data=df, family=binomial)
summary(var12_model)
```

### var13

```{r}
table(df$var13)
```

```{r}
var13_model <- glm(formula = disease ~ var13, data = df, family = binomial)
summary(var13_model)
```

### var10

```{r}
hist_by_lc <- function(covariate){
  distribution <- df %>%
    select({{ covariate }}, disease) %>% 
    filter(!is.na(disease))
  hist <- distribution %>% 
    ggplot(aes(x = {{ covariate }}, fill = disease)) +
            geom_histogram() +
            facet_wrap(~disease, scales = "free")
  print(hist)
}
```

```{r}
hist_by_lc(`var10`)
```

```{r}
var10_test <- t.test(var10 ~ disease, data = df, var.equal = FALSE)
print(var10_test)
```

### var11

```{r}
hist_by_lc(var11)
```


### log regression of PC1, PC2 and PC3

```{r}
pc_model <- glm(disease ~ pc1 + pc2 + pc3, data = df, family = binomial)
summary(pc_model)
```

# Step 2: First model

```{r}
first_model <- glm(disease ~ pc1 + pc2 + pc3 + var10 + var11 + var1 + var2_1 + var2_2 + var2_3 + var2_4 + var4 + `var5_1` + var5_2 + `var5_3 + var6_1 + var6_2 + var7 + var14 + var12 + var13,
                   data=df,
                   family=binomial)
summary(first_model)
```

### merging var 3 categories

```{r}
df <- df %>% 
  dummy_cols(select_columns = "var3", ignore_na = TRUE)
```

```{r}
first_model_ethnicity <- glm(disease ~ pc1 + pc2 + pc3 + var10 + var11 + var1_Female+ var6_Yes + var14 + var12 + var13,
                   data=df,
                   family=binomial)
summary(first_model_ethnicity)
```

## Model with covariates of p\<0.05

### coefficient changes \>20% : none

```{r}
second_model <- glm(disease ~ pc1+ var10 + var11 + var1_Female + var14+ var12 + var13,
                   data=df,
                   family=binomial)
summary(second_model)
```

# Step 3: mitigate coefficient changes

# Step 4: returning variables excluded in step 1

```{r}
step_4_model <- glm(disease ~ pc1 + var10 + var11 + var1_Female + var14+ var12 + var13 + var8,
                   data=df,
                   family=binomial)
summary(step_4_model)
```

```{r}
step_4_model <- glm(disease ~ pc1 + var10 + var11 + var1_Female + var14+ var12 + var13+ cardiovascular_heart.disease,
                   data=df,
                   family=binomial)
summary(step_4_model)
```

```{r}
step_4_model <- glm(disease ~ pc1 + var10 + var11 + var1_Female + var14+ var12 + var13+ var9,
                   data=df,
                   family=binomial)
summary(step_4_model)
```

```{r}
step_4_model <- glm(disease ~pc1 + var10 + var11 + var1_Female + var14+ var12 + var13+ diabetes,
                   data=df,
                   family=binomial)
summary(step_4_model)
```

```{r}
step_4_model <- glm(disease ~ pc1 + var10 + var11 + var1_Female + var14+ var12 + var13+ hypertension,
                   data=df,
                   family=binomial)
summary(step_4_model)
```

# Preliminary Effects Model:

```{r}
summary(second_model)
```

# Step 5: Assessing linearity

## Obtain estimated log odds

```{r}
yhat <- predict(second_model, type = "response", newdata = df)
df$yhat <- yhat
df_clean <- df %>% 
  filter(!is.na(pc1)) %>% 
  filter(!is.na(disease)) %>% 
  filter(!is.na(yhat)) %>% 
  filter(!is.na(var10)) %>% 
  filter(!is.na(var11)) %>% 
  fiter(!is.na(var12)) %>% 
  filter(!is.na(var13)) %>% 
  mutate(logit_p = log((yhat)/(1 - yhat)))
```

## pc1

### lowess plot

```{r}
plot_lowess <- function(x){
  lowess_df_clean <- df_clean %>% 
    filter(!is.na({{x}}))
  plot(df_clean[[x]], df_clean$logit_p, main = "LOWESS", xlab = {{x}}, ylab = "Log Odds of Long Covid")
  lowess_curve <- lowess(df_clean[[x]], df_clean$logit_p)
  lines(lowess_curve, col = "red", lwd = 2)
}
plot_lowess("pc1")
```

## pc3

```{r}
plot_lowess("pc2")
```

## var10

```{r}
plot_lowess("var10")
```

## var11

```{r}
plot_lowess("var11")
```

## var12

```{r}
plot_lowess("var12")
```

## var13

```{r}
plot_lowess("var13")
```

# Main Effects Model:

```{r}
summary(second_model)
```

# Step 6: Checking interactions

## create interaction terms

```{r}
print(mean(df_clean$var10))
print(mean(df_clean$var13))
print(mean(df_clean$var12))
print(mean(df_clean$var11))
print(mean(df_clean$pc1))
print(mean(df_clean$pc2))
```

```{r}
df_inter <- df_clean %>% 
  mutate(
    ctr_var10 = (var10 - 52.27096),
    ctr_var12 = (var12 - 2.33607),
    ctr_var13 = (var13 - 0.4961915),
    ctr_pc1 = (pc1 - 3.645434),
    ctr_pc2 = (pc2 - 0.3261621),
    ctr_var11 = (var11 - 27.51),
    var10_pc1 = ctr_pc1*ctr_var10,
    var10_pc2 = ctr_pc2*ctr_var10,
    var12_var13 = ctr_var12*ctr_var13,
    var10_var12 = ctr_var10*ctr_vacc
  )
```

## testing each interaction

```{r}
interaction_test <- function(b1, b2 = NULL, b3 = NULL, b4 = NULL) {
interaction_formula <- paste("disease ~ pc1 + var10 + var11 + var1 + var14+ var12 + var13")
if (!missing(b1)) interaction_formula <- paste(interaction_formula, "+", b1)
if (!missing(b2)) interaction_formula <- paste(interaction_formula, "+", b2)
if (!missing(b3)) interaction_formula <- paste(interaction_formula, "+", b3)
if (!missing(b4)) interaction_formula <- paste(interaction_formula, "+", b4)
interaction_model <- glm(as.formula(interaction_formula),
                   data=df_inter,
                   family=binomial)
reduced_model <- glm(disease ~ pc1 + var10 + var11 + var1_Female + var14+ var12 + var13,
                   data=df,
                   family=binomial)
lrt <- anova(reduced_model, interaction_model, test = "LRT")
ifelse(lrt[,5] < 0.05, print(lrt), print(NA))
}
```

```{r}
interaction_test("var12_var13")
interaction_test("var10_var12")
```

# Preliminary Final Model:

```{r}
df_inter_clean <- df_inter %>% 
  filter(!is.na(var1)) %>% 
  filter(!is.na(var4)) %>% 
  filter(!is.na(var3)) %>% 
  filter(!is.na(var7)) %>% 
  filter(!is.na(var14))
set.seed(32)
no_lc <- df_inter_clean%>% 
  filter(disease==0)
set <- sample(nrow(no_lc), nrow(no_lc)*0.95)
df_balance <- df_inter_clean
df_balance <- df_balance[-set,]
table(df_balance$disease)
```

```{r}
prelim_final <- glm(disease ~pc1+ctr_var10+ var11 + var1+ctr_var12 + ctr_var13+var12_var13+var10_var12,
                   data=df_inter,
                   family=binomial)
summary(prelim_final)
```

# Step 7: Assess fit using HL test

```{r}
yhat_final <- predict(prelim_final, type = "response")
hl_test <- hoslem.test(df_inter$disease, yhat_final, g = 348)
print(hl_test)
```

# Export Results

```{r}
final <- prelim_final
rtf_or <- RTF("df_OR.rtf")
addParagraph(rtf_or, "Table 3: Univariate logistics analysis of adjusted odds ratios for the effect of PC1, PC2 and PC3.")
addTable(rtf_or, as.data.frame(cbind(round(exp(coefficients(final)), 2), round(exp(confint(final, test = "Rao")), 2), round(coef(summary(final))[,4], 3))), row.names = TRUE)
done(rtf_or)
```

```{r}
rtf_pc.or <- RTF("pc_OR.rtf")
addTable.RTF(rtf_pc.or, as.data.frame(cbind(round(exp(coefficients(pc_model)), 2), round(exp(confint(pc_model, test = "Rao")), 2), round(coef(summary(pc_model))[,4], 3))), row.names = TRUE)
done(rtf_pc.or)
```

```{r}
print(confint(final, test = "Rao"))
```
